% background.tex

\chapter{Background}

\section{History and current situation}%
\label{sec:history_and_current_situation}
The basic idea of NTP was developed in the early 1980s. NTP version 0 was
implemented in 1985, and it is documented in RFC~958~\cite{rfc958}. 
% But the actual implementation did a little bit more. 
The NTP packet header and the calculations of offset and delay (as will be
discussed later) from that NTP version are still being used today. At that
time, the nominal accuracy could be in the low tens of milliseconds on an
Ethernet network and better than 100 milliseconds on paths spanning the
Atlantic Ocean.

NTP version 1 was documented in RFC~1059~\cite{rfc1059} one year later. It
contains a specification of the protocol and algorithms.

The NTP version 2 specification was documented in RFC~1119~\cite{rfc1119} in 1989.
This version
includes new protocols for use in managing NTP servers and clients and a
cryptographic authentication scheme.
At about the same time, another time synchronization protocol was presented,
known as the Digital Time Synchronization Service (DTSS)~\cite{DTSS}.

The NTP version 3 specification appeared in 1992, following
RFC~1305~\cite{rfc1305}. This version combined the good ideas from both NTP and
DTSS\null.

After that, the Simple Network Time Protocol version 4 was documented in
RFC~2030~\cite{rfc2030}
in 1996. SNTP is used if users do not need the ultimate performance of NTP or
it is not otherwise justified.

The latest version of NTP (at the time of writing) follows
RFC~5905~\cite{rfc5905}, which is NTP version 4; this was presented in 2010.
% history.pdf

\section{Reference clocks}%
\label{sec:reference_clocks}
``A reference clock is some device or machinery that spits out the current
time''~\cite{reference_clock}.
% http://www.ntp.org/ntpfaq/NTP-s-algo.htm#Q-REFCLK
The most important aspect of a reference clock is its accuracy; it must be
synchronized to some time standard (usually UTC\null). If we want to
synchronize a device to UTC, we require reference clocks as sources which are
themselves synchronized to UTC\null. Some of them are
available from government dissemination services~\cite{redbook}, including:
\begin{itemize}
    \item Global Positioning System (GPS) and Long-Range navigation (LORAN-C)
        systems
    \item WWV/H and WWVB radio time/frequency stations
    \item U.S. Naval Observatory (USNO) and National Institutes of Science and
        Technology (NIST\null, formerly the National Bureau of Standards [NBS])
        telephone modem services in the United States
    \item Similar systems and services in other countries
\end{itemize}
Generally, we want all devices synchronized to UTC as well as reference clocks.
But in some cases, we may just want all devices in a local area network
synchronized to a certain ``standard'' time such as the system time of a
certain device among them. In this case, no UTC reference clock is in the
subnet. However, we can consider the device to which others are synchronized as
a reference clock. 

\section{Adjusting system time}%
\label{sec:adjusting_system_time}
As mentioned in Section~\ref{sec:computers_clocks}, every computer should have
a device outputting a signal with constant frequency to increase the number in
a counter, which is used to calculate system time. However, in practice, the
frequency may be neither accurate nor consistent. In this case, even if at some
time we synchronized the system time to UTC or an alternative reference clock,
it will be inaccurate later. This fact makes it necessary to synchronize a
device frequently.

When the system clock is adjusted by NTP\null, both the time and the frequency
will be adjusted. For Unix systems, NTP uses the \verb|ntp_adjtime()| or
the \verb|adjtime()| system call to adjust the system time when the offset from
the reference time is small.
The benefit of using these system calls is that they will synchronize the
system time step by
step over a period of time, adjusting only a small amount in one step, and finally
synchronize the system clock to the reference clock. In this case, any running
program which is sensitive to
time, such as a logging program, will only be slightly impacted. If we want to
change system time backward some amount, it will not make later things
apparently happen earlier in the log. If we want to change system time forward,
there will not be a gap in the time line, where nothing happened during that
time. However, if the offset from the reference time is large, NTP may make a
large, abrupt change in the system time.

In next chapter, we give an overview of the architecture of the network
operated by NTP and the internal processes of a NTP daemon program.

\section{NTP daemon and query program}%
\label{sec:ntp_daemon_and_query_program}
There is an NTP daemon program, called \verb|ntpd|, which is an implementation of
NTP\null. The \verb|ntpd| daemon program synchronizes the system clock. The
query program \verb|ntpq| allows users to get some information about the running
status of the daemon program, including server information,
statistics and kernel status.

Here is an example which shows the output of the \verb|kern| command of the
\verb|ntpq| program. In this example, we can see that the estimated error of
the system clock is 1.393~ms. The meaning of some of the other output lines
will be touched on in later chapters.
\begin{verbatim}
ntpq> kern
associd=0 status=0638 leap_none, sync_ntp, 3 events, no_sys_peer,
pll offset: -0.235878
pll frequency: -44.3577
maximum error: 728.128
estimated error: 1.393
kernel status: pll nano mode=fll
pll time constant: 10
precision: 0.001
frequency tolerance: 500
pps frequency: 0
pps stability: 0
pps jitter: 0
calibration interval 0
calibration cycles: 0
jitter exceeded: 0
stability exceeded: 0
calibration errors: 0
\end{verbatim}

% \section{Concepts}%
% \label{sec:concepts}

% (Note: should this part be here or somewhere else or no need?)


