% overview.tex

\chapter{Overview}
% Three kinds of devices.
NTP operates on three different kinds of devices: primary servers, secondary
servers and clients. Primary servers are directly synchronized to reference
clocks and provides synchronization service to other devices.  Secondary severs
are synchronized to other servers instead of reference clocks and also provide
synchronization services. Clients are synchronized to other servers but they do
not provide synchronization to others~\cite{rfc5905}.

\section{Protocol modes}%
\label{sec:protocol_modes}
NTP has three basic protocol modes: \emph{server/client}, \emph{symmetric}
and \emph{broadcast}. These protocol modes are also called association modes. 
%(NOTE: I DID NOT FIND ANY DEFINITION OF ASSOCIATION!)
In the server/client variant, clients send requests to a server for
synchronization, and servers handle requests and send responses. We say
``clients pull synchronization from servers''. The broadcast mode is somewhat
the opposite of server/client mode; in this case we say ``servers push
synchronization to clients''. Symmetric mode is a combination of two
server/client modes and two broadcast modes. In the symmetric variant, a peer
operates as both client and server and each peer both pushes and pulls
synchronization to and from each other~\cite{rfc5905}.

There are two interleaved modes which we will discuss in
Section~\ref{sub:timestamps_in_ntp_packet}.

\section{Layered Network}
\label{sec:Layered_network}
The NTP network or subnet has a hierarchy architecture, which is divided into
several levels. Each level is called a stratum. All primary servers are in
stratum one, which is the highest stratum. Secondary servers are assigned the
stratum number one greater than stratum number of the server to which it
synchronizes to. If a secondary server synchronizes to a server in stratum 3,
it is in stratum 4. The maximum stratum number is 16~\cite{rfc5905}.
Typically, clients do not have stratum number, but we can treat them as being
in a stratum by using the same rule as secondary servers.  As the stratum
number increases, the accuracy of time estimates decreases.

Figure~\ref{fig:layered_network} shows the basic structure of NTP subnets. The
arrows indicate the direction of synchronization. An arrow from Server~3 to
Server~1 indicates that Server~3 communicates with Server~1 and tries to
synchronize to it, but Server~3 may not actually synchronize to Server~1.

% This figure is made by myself, may need improvement
% fig:layered_network
\input{figures/overview_layer}

% Feb 20 2019, Dr. Diamond gave some advise about this paragraph
As people want their devices to have accurate time, it is better to synchronize to
primary servers. But this can make primary servers very busy. Based on the rule
of engagement for NTP users, ``clients should avoid using primary servers
whenever possible''. Even though, some busy primary servers have more than 700
clients~\cite{rules}. 
% http://support.ntp.org/bin/view/Servers/RulesOfEngagement
It is said that the synchronizations provided by secondary servers in stratum 2
are accurate enough for most end-users. Good practices suggest that clients
should use three to five servers~\ref{rules}. 
In practice, multiple NTP servers are organized into a group known as a
\emph{pool}. Each such pool has its own DNS entry, and clients wishing to
synchronize to one or more servers in a particular pool use the pool's DNS
name. When a client tries to connect to the pool, ``DNS round robin" is used to
select a random server from the pool. The client then connects to this randomly
selected server~\cite{pool}.

% In practice, there are some pools
% of secondary servers.  Clients can assign a pool as a server which they want to
% synchronize to. When the pool receive the connection request, it uses DNS round
% robin to select a random server from the pool. Then client actually connects to
% the selected server~\cite{pool}.
% http://support.ntp.org/bin/view/Servers/NTPPoolServers

\section{Processes}
\label{sec:processes}
% Architecture of client processes.
As seen in Figure~\ref{fig:architecture_overview} (which is an adaptation
of a figure in~\cite{redbook}), the operations of an NTP client
can be separated into the following processes:
\begin{itemize}
    \item Peer/Poll Processes\\
        This part deals with communications between client and servers. It has
        poll processes and peer processes; poll processes send requests to
        servers, while peer processes keep track of a set of statistics for every
        server.
    \item System Process\\
        The system process has algorithms to do filtering and correction work
        to the statistics of all peers, then passes them to clock discipline
        process.
    \item Clock Discipline Process\\
        The clock discipline process is like a lowpass filter to smooth the
        data.  This process also maintains the poll intervals~\cite{redbook}.
        % red book p. 40
    \item Clock Adjust Process\\
        This part adjusts the system clock to make it continuous and monotonic
        approximately which means that the system clock will not go
        backward~\cite{redbook}. 
        % red book p. 40
\end{itemize}

% This figure is made based on the figure from
% https://www.eecis.udel.edu/~mills/ntp/html/warp.html
% fig:architecture_overview
\input{figures/overview_arch}

The structure of NTP is like a loop. In short, it requests synchronization
information from servers, and after getting one or more respones, it processes
the received information and (if necessary) adjust the system time.  After
that, feedback from the clock adjust process is used to control poll processes.

In later chapters, we will discuss the peer/poll processes and the system
process separately, then discuss the clock discipline process and the clock
adjust process together. Note that, in official NTP documents, as well as this
thesis, ``server'' and ``peer'' are used interchangeably.

% (What is the stuff I want to add here when I was taking shower on Fer 5?)

